---
title: "Reproducible-research-project1"
---
## Introduction
It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the “quantified self” movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

## Data
The data for this assignment was downloaded from the course web
site:

* Dataset: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) [52K]

The variables included in this dataset are:

* steps: Number of steps taking in a 5-minute interval (missing
    values are coded as `NA`)

* date: The date on which the measurement was taken in YYYY-MM-DD
    format

* interval: Identifier for the 5-minute interval in which
    measurement was taken



## Loading and preprocessing the data

Load and preprocessing data. Ignore missing values.
```{r}
setwd("~/Desktop/Reproducible-Research")
dat <- read.csv("activity.csv",header = TRUE)
dat$date <- as.Date(dat$date) 
dat.ig.na <- na.omit(dat) 
```


## What is mean total number of steps taken per day?
Sum steps by day, create Histogram, and calculate mean and median.
```{r} 
dailysteps <- rowsum(dat.ig.na$steps, format(dat.ig.na$dat, '%Y-%m-%d')) 
dailysteps <- data.frame(dailysteps) 
names(dailysteps) <- ("steps") 

hist(dailysteps$steps, main = "Total number of steps taken per day ", xlab="Number of Steps")
rmean <- mean(dailysteps$steps)
rmedian <- median(dailysteps$steps)
```

The `mean` total number of steps taken per day is `r rmean` and the `median` is `r rmedian`.

## What is the average daily activity pattern?

```{r}
steps_by_interval <- aggregate(steps ~ interval, dat, mean)

plot(steps_by_interval$interval,steps_by_interval$steps, type="l", xlab="Interval", ylab="Number of Steps",main="Average Number of Steps per Day by Interval")

max_interval <- steps_by_interval[which.max(steps_by_interval$steps),1]
```

The 5-minute interval, on average across all the days in the data set, containing the maximum number of steps is `r max_interval`.

## Impute missing values. Compare imputed to non-imputed data.
Note that there are a number of days/intervals where there are missing values (coded as 𝙽𝙰). The presence of missing days may introduce bias into some calculations or summaries of the data.

Our statage is replace the missing value by inserting the average for each interval.
```{r}
incomplete <- sum(!complete.cases(dat))
imputed_data <- transform(dat, steps = ifelse(is.na(dat$steps), steps_by_interval$steps[match(dat$interval, steps_by_interval$interval)], dat$steps))
```


```{r}
imputed_data[as.character(imputed_data$date) == "2012-10-01", 1] <- 0
```

Recount total steps by day and create Histogram. 
```{r}
new_steps_by_day <- aggregate(steps ~ date, imputed_data, sum)
hist(new_steps_by_day$steps,main = "Total number of steps taken per day", col="blue", xlab="Number of Steps")

#Create Histogram to show difference. 
hist(dailysteps$steps, main = "Total number of steps taken per day", col="red", xlab="Number of Steps", add=T)
legend("topright", c("Imputed", "Non-imputed"), col=c("blue", "red"), lwd=10)
```

Calculate new mean and median for imputed data. 
```{r}
rmean.i <- mean(new_steps_by_day$steps, na.rm=TRUE)
rmedian.i <- median(new_steps_by_day$steps, na.rm=TRUE)
```

Calculate difference between imputed and non-imputed data.
```{r}
mean_diff <- rmean.i - rmean
med_diff <- rmedian.i - rmedian
```

Calculate total difference.
```{r}
total_diff <- sum(new_steps_by_day$steps) - sum(dailysteps$steps)
```
* The imputed data mean is `r rmean.i`
* The imputed data median is `r rmedian.i`
* The difference between the non-imputed mean and imputed mean is `r mean_diff`
* The difference between the non-imputed mean and imputed mean is `r med_diff`
* The difference between total number of steps between imputed and non-imputed data is `r total_diff`. Thus, there were `r total_diff` more steps in the imputed data.


## Are there differences in activity patterns between weekdays and weekends?
Created a plot to compare and contrast number of steps between the week and weekend. There is a higher peak earlier on weekdays, and more overall activity on weekends.  
``` {r}
weekdays <- c("Monday", "Tuesday", "Wednesday", "Thursday", 
              "Friday")
imputed_data$dow = as.factor(ifelse(is.element(weekdays(as.Date(imputed_data$date)),weekdays), "Weekday", "Weekend"))

steps_by_interval_i <- aggregate(steps ~ interval + dow, imputed_data, mean)

library(lattice)

xyplot(steps_by_interval_i$steps ~ steps_by_interval_i$interval|steps_by_interval_i$dow, main="Average Steps per Day by Interval",xlab="Interval", ylab="Steps",layout=c(1,2), type="l")

```